---
layout: post
title: "Titanic Survival Analysis"
author: "Payam Bahreyni"
date: 2016-03-22
categories: [tutorials]
tags : [machine learning]
---

## Loading Data & Initial Analysis

```{r package_options, include=FALSE}
knitr::opts_knit$set(root.dir= "~/Documents/Projects/Data Camp - Titanic")
```

```{r, message=FALSE, warning=FALSE}
source("library.R")
```

```{r}
# Load data
train.data <- read.csv("input/train.csv", stringsAsFactors = F)
test.data <- read.csv("input/test.csv", stringsAsFactors = F)

train.outcome <- train.data$Survived
train.outcome <- factor(train.outcome)

# Remove "Survived" column
train.data <- train.data[,-2]

train.len <- nrow(train.data)
test.len <- nrow(test.data)

# Combine training & testing data
full.data <- rbind(train.data, test.data)

# Turn categorical vars into factor
full.data$Pclass <- factor(full.data$Pclass)
full.data$Sex <- factor(full.data$Sex)
full.data$Embarked <- factor(full.data$Embarked)

str(full.data)

# Split data back into training and testing
train.data <- full.data[1:train.len, ]
test.data <- full.data[(train.len+1):nrow(full.data), ]

sapply(train.data[, -c(1, 3, 8, 10)], summary)

train.data$Survived <- train.outcome
```

### Findings 

* Most people are travelled in Class 3.
* Most passengers were male.
* For the passengers with Age specified, 50% are older than 28. We have a lot of missing values in Age variable.
* Most passengers did not travel with their spouse or siblings on board. 
* Most passengers do not have their parents and children on board.
* The 75% of the passengers paid less than $31 for fare, and the maximum fare paid was $512.
* Most passengers came on board on the port specified by "S".

## EDA

Let's take a look at different groups and their survival rate.

```{r}
prop.table(table(train.data$Sex, useNA = "ifany"))
prop.table(table(train.data$Pclass, useNA = "ifany"))
```

Majority of passengers are men (65%) and passengers have 3 different classes, 1 (24%), 2 (21%), or 3 (55%).

### Survival Rate

```{r}
table(train.data$Survived, useNA = "ifany")
prop.table(table(train.data$Survived, useNA = "ifany"))

table(train.data$Sex, train.data$Survived)
prop.table(table(train.data$Sex, train.data$Survived), margin = 1)
prop.table(table(train.data$Pclass, train.data$Survived), margin = 1)
```

Most people didn't survive (62%). But the survival rate is not the same across different groups. Females had higher chance of survival, 74% as compared to 19% for men. Class 1 passengers had 63% chance of survival, compared to 47% and 24% for class 2 and 3, respectively.

```{r}
prop.table(table(train.data$Pclass, train.data$Sex), margin = 1)
prop.table(table(train.data$Pclass, train.data$Sex, train.data$Survived), margin = 3)
```

The survival rate in different classes may have some relationship to percentage of women in those classes.

```{r}
round(train.len * 100/(train.len + test.len), digits = 2)
round(test.len * 100/(train.len + test.len), digits = 2)
```

The data is partitioned into 68% for training dataset and 32% for test data set.

## Graphical Analysis
```{r}
# Graphical
library(ggplot2)
library(scales)

ggplot(train.data, aes(Sex, fill= Survived)) +
    geom_bar(stat = "bin", position = "stack") +
#    geom_text(stat = "bin", aes(label= ..count.., vjust= 1),
#              position = position_stack()) +
    ggtitle("Women have higher chance of survival")

ggplot(train.data, aes(x=Pclass, fill= Survived)) +
    geom_bar(stat = "bin", position = "stack") +
    ggtitle("In class 1 most people survived and in class 3 most did not")

ggplot(train.data, aes(x=Pclass, fill= Survived))+
    geom_bar(stat = "bin", position = "stack") +
    facet_wrap(~Sex) +
    ggtitle("Most women survive across passenger classes")

ggplot(train.data, aes(y=Fare, x= Survived)) +
    geom_boxplot() +
    facet_wrap(~Pclass, ncol = 3, scales = "free") +
    ggtitle("People who survived paid higher fare on average")
```

## Statistical Analysis

We saw that fare may have some effect on the survival rate. Let's see if the effect is real.

```{r}
aggregate(Fare ~ Survived + Pclass, data=train.data, mean)

with(train.data, t.test(Fare[Pclass == 1 & Survived == 1], 
                        Fare[Pclass == 1 & Survived == 0]))

with(train.data, t.test(Fare[Pclass == 2 & Survived == 1], 
                        Fare[Pclass == 2 & Survived == 0]))

with(train.data, t.test(Fare[Pclass == 3 & Survived == 1], 
                        Fare[Pclass == 3 & Survived == 0]))
```

So, we see that in class 1, the difference in fare between passengers who survived and who didn't is statistiscally significant. Maybe the rich found a way to buy lifeboats :).

## Machine Learning

Let's do some modeling to see how we can predict the survival rate given other variables about the passengers.
```{r}
library(caret)
library(rpart)

sex.model.tree <- train(Survived ~ Sex + Pclass, data= train.data, method= "rpart")
sex.class.survival <- predict(sex.model.tree, train.data)
confusionMatrix(sex.class.survival, train.data$Survived)

plot.decision.tree(sex.model.tree$finalModel)
```

Although `Sex` and `Pclass` looked like good predictors for `Survival`, `Pclass` didn't get picked up by `rpart` as it didn't add much information to the decision tree.

```{r}
sex.class.fare.tree <- train(Survived ~ Sex + Pclass + Fare, data= train.data, method= "rpart")
sex.class.fare.survival <- predict(sex.class.fare.tree, train.data)
confusionMatrix(sex.class.fare.survival, train.data$Survived)

plot.decision.tree(sex.class.fare.tree$finalModel)
```

As expected, a mix of passenger class and fare plays a role in our prediction this time. We used to predict that women survive with a 74% chance. Now, we can be more accurate. Women have a 95% chance of survival if not in class 3. From class 3 women, 11% of people who paid fare of more than $23, and 59% of people who paid less survived. Let's see if it makes sense.

```{r}
ggplot(train.data[train.data$Pclass== "3" & train.data$Sex == "female",], aes(x= Survived, y= Fare)) +
    geom_boxplot() +
    ggtitle("Class 3 women who survived paid less fare on average")

without.age.tree <- train(Survived ~ Sex + Pclass + Fare + SibSp + Parch + Embarked,
                          data= train.data,
                          method= "rpart")
plot.decision.tree(without.age.tree$finalModel)
```

Adding all the variables but age didn't change the decision tree.

### Method 2: Logistic Regression

```{r}
sex.class.fare.logit <- glm(Survived ~ Sex + Pclass + Fare, 
                            family = "binomial", data= train.data)
sex.class.fare.survival <- predict(sex.class.fare.logit, train.data, type = "response")
head(sex.class.fare.survival)
summary(sex.class.fare.logit)

without.age.logit <- glm(Survived ~ Sex + Pclass + Fare + SibSp + Parch + Embarked, 
                            family = "binomial", data= train.data)
without.age.survival <- predict(without.age.logit, train.data, type = "response")
head(without.age.survival)
summary(without.age.logit)

```

## Feature Creation
```{r, eval= FALSE}
# Create the column child, and indicate whether child or no child
train2 <- train.data
train2$Child <- NA
train2$Child <- ifelse(is.na(train2$Age), NA, ifelse(train2$Age < 18, 1, 0))
train2$Child <- factor(train2$Child)

summary(train2$Age)
round(sum(is.na(train2$Age))* 100/nrow(train2), digits = 2)

train2$AgeGroup <- cut(train2$Age, breaks=c(0, 5, 10, 18, 28, 38, 80),
                       labels=c("<5", "5-10", "10-18", "18-28", "28-38", ">38"))
prop.table(table(train2$AgeGroup, useNA = "ifany"))

# Two-way comparison
prop.table(table(train2$Child, train2$Survived), margin= 1)
# Result: There is a difference in survival rate whether child or not

ggplot(train2, aes(x=Child, fill= Survived)) +
    geom_bar(stat = "bin", position = "dodge")

ggplot(train2, aes(x= Sex, y= AgeGroup, color= Survived)) +
    geom_jitter()

prop.table(table(train2$Sex, train2$AgeGroup), margin = 1)

aggregate(Survived ~ Sex + AgeGroup, data= train2, function(x) {
    sum(x$Survived == "1")/nrow(x)
})
```

